Plug-and-play dual-tree algorithm runtime analysis

### 2015-01-08 bounds-files/imbalance.tex
updated abstract and intro, typo fixes, added alternative imbalance notes
Minor improvements after a read-through.  I need to fix the bibliography (UGH) but this is all that I think is necessary in terms of changes to the paper.

It is well-known that imbalance in trees leads to degradation in performance;
for instance, a $kd$-tree node with every descendant except one in its left
child is effectively useless.  A $kd$-tree full of nodes like this will perform
abysmally for nearest neighbor search, and it is not hard to generate a
pathological dataset that will cause a $kd$-tree of this sort.

@@

An imbalanced cover tree can happen in practice, and in the worst cases, the
imbalance may be far worse than the simple graphs of Figure \ref{fig:imbalance}.
Consider a dataset with a single outlier which is very far away from all of the
other points\footnote{Note also that for an outlier sufficiently far away, the
expansion constant is $N - 1$.}.  Figure \ref{fig:outlier} shows what happens in
this situation: the root node has two children; one of these children has only
the outlier as a descendant, and the other child has the rest of the points in
the dataset as a descendant.  In fact, it is easy to find datasets with a
handful of outliers that give rise to a chain-like structure at the top of the
tree: see Figure \ref{fig:outliers} for an illustration\footnote{As a side note,
this behavior is not limited to cover trees, and can happen to mean-split
$kd$-trees too, especially in higher dimensions.}.

@@

A tree that has this chain-like structure all the way down is going to perform
horrendously; motivated by this observation, we define a measure of tree
imbalance.

@@

\begin{equation}
i_n(\mathscr{N}_i) = \begin{dcases*}
  s_p - s_i - 1 & if $\mathscr{N}_i$ is not a leaf and not the root node \\
  \max(s_p - s_{\min} - 1, \; 0) & if $\mathscr{N}_i$ is a leaf \\
  0 & if $\mathscr{N}_i$ is the root node.
  \end{dcases*}
\end{equation}
\end{defn}

@@

A perfectly balanced cover tree $\mathscr{T}_b$ with no missing levels has
imbalance $i_t(\mathscr{T}_b) = 0$ (for instance, Figure
\ref{fig:imbalance-good}).  A worst-case cover tree $\mathscr{T}_w$ which is
entirely a chain-like structure with maximum scale $s_{\max}$ and minimum scale
$s_{\min}$ will have imbalance $i_t(\mathscr{T}_w) \sim N (s_{\max} -
s_{\min})$.  Because of this chain-like structure, each level has only one node
and thus there are at least $N$ levels; or, $s_{\max} - s_{\min} \ge N$, meaning
that in the worst case the imbalance is quadratic in $N$.

However, for most real-world datasets with the cover tree implementation in {\bf
mlpack} \citep{mlpack2013} and the reference implementation
\citep{langford2006}, the tree imbalance is near-linear with the number of
points.  Most of the cover tree imbalance is contributed by leaf nodes whose
parent has scale greater than $s_{\min}$.  At this time, no cover tree
construction algorithm specifically aims to minimize imbalance.

>>>>

It is well-known that imbalance in trees leads to degradation in performance;
for instance, a $kd$-tree node with every descendant in its left child except
one is effectively useless.  A $kd$-tree full of nodes like this will perform
abysmally for nearest neighbor search, and it is not hard to generate a
pathological dataset that will cause a $kd$-tree of this sort.

@@

An imbalanced cover tree can happen in practice, and in the worst cases, the
imbalance may be far worse than the simple graphs of Figure \ref{fig:imbalance}.
Consider a dataset with a single outlier which is very far away from all of the
other points\footnote{Note also that for an outlier sufficiently far away, the
expansion constant is $N - 1$, so we should expect poor performance with the
cover tree anyway.}.  Figure \ref{fig:outlier} shows what happens in
this situation: the root node has two children; one of these children has only
the outlier as a descendant, and the other child has the rest of the points in
the dataset as a descendant.  In fact, it is easy to find datasets with a
handful of outliers that give rise to a chain-like structure at the top of the
tree: see Figure \ref{fig:outliers} for an illustration\footnote{As a side note,
this behavior is not limited to cover trees, and can happen to mean-split
$kd$-trees too, especially in higher dimensions.}.

@@

A tree that has this chain-like structure all the way down, which is similar to
the $kd$-tree example at the beginning of this section, is going to perform
horrendously; motivated by this observation, we define a measure of tree
imbalance.

@@

\begin{equation}
i_n(\mathscr{N}_i) = \begin{dcases*}
  s_p - s_i - 1 & if $\mathscr{N}_i$ is not a leaf and not the root node \\
  \max(s_p - s_{\min} - 1, \; 0) & if $\mathscr{N}_i$ is a leaf \\
  0 & if $\mathscr{N}_i$ is the root node.
  \end{dcases*}
  \label{eqn_node_imbalance}
\end{equation}
\end{defn}

@@

A perfectly balanced cover tree $\mathscr{T}_b$ with no missing levels has
imbalance $i_t(\mathscr{T}_b) = 0$ (for instance, Figure
\ref{fig:imbalance-good}).  A worst-case cover tree $\mathscr{T}_w$ which is
entirely a chain-like structure with maximum scale $s_{\max}$ and minimum scale
$s_{\min}$ will have imbalance $i_t(\mathscr{T}_w) \sim N (s_{\max} -
s_{\min})$.  Because of this chain-like structure, each level has only one node
and thus there are at least $N$ levels; or, $s_{\max} - s_{\min} \ge N$, meaning
that in the worst case the imbalance is quadratic in $N$\footnote{Note that in
this situation, $c \sim N$ also.}.

However, for most real-world datasets with the cover tree implementation in {\bf
mlpack} \citep{mlpack2013} and the reference implementation
\citep{langford2006}, the tree imbalance is near-linear with the number of
points.  Generally, most of the cover tree imbalance is contributed by leaf
nodes whose parent has scale greater than $s_{\min}$.  At this time, no cover
tree construction algorithm specifically aims to minimize imbalance.

### 2015-01-08 bounds-files/sec01-supporting-lemmas.tex
Minor improvements after a read-through.  I need to fix the bibliography (UGH) but this is all that I think is necessary in terms of changes to the paper.

For simplicity, the algorithms considered in this paper will be presented in a
tree-independent context, as in \citet{curtin2013tree}, but the only type of
tree we will consider is the cover tree \citep{langford2006}, and the only type
of traversal we will consider is the cover tree pruning dual-tree traversal,
which we will describe below.

@@

A consequence of this definition is if there exists a node $\mathscr{N}_i$,
containing the point $p_i$ at some scale $s_i$, then there will also exist a
self-child node $\mathscr{N}_{ic}$ containing the point $p_i$ at scale $s_i - 1$
which is a child of $\mathscr{N}_i$.  In addition, every descendant point of the
node $\mathscr{N}_i$ is contained within a ball of radius $2^{s_i + 1}$ centered
at the point $p_i$; therefore, $\lambda_i = 2^{s_i + 1}$ and $\mu_i = p_i$
(Table \ref{tab:notation}).

Note that the cover tree may be interpreted as an infinite-leveled tree, with
$C_{\infty}$ containing only the root point, $C_{-\infty} = S$, and all levels
between defined according to the definitions above.  \citep{langford2006} find
this representation (which they call the {\it implicit} representation) easier
for description of their algorithms and some of their proofs.  But naturally,
this is not suitable for implementation; hence, there is an {\it explicit}
representation in which all nodes that have only a self-child are coalesced.

In our work here, we consider only the explicit representation of a cover tree,
and do not concern ourselves with the construction of the tree\footnote{A batch
construction algorithm is given by \citet{langford2006}, called
\texttt{Construct}.}.

@@

Take $f(x)$ as a uniform spherical distribution in $\mathcal{R}^d$: for any $|x|
\le 1$, $f(x)$ is a constant; for $|x| > 1$, $f(x) = 0$.  It is easy to see that
$c_f$ in this situation is $2^d$. % To demonstrate this convergence, Table
%\ref{tab:ec_scaling} shows results for 10 trials of empirically calculated
%expansion constants for datasets drawn from this distribution, for varying $d$.

@@

There are some other important observations about the behavior of $c$.  Adding a
single point to $S$ may increase $c$ arbitrarily: consider a set $S$ distributed
entirely on the surface of a unit hypersphere.  If one adds a single point at
the origin, producing the set $S'$, then $c$ explodes to $|S'|$ whereas before
it may have been much smaller than $|S|$.  Adding a single point may also
decrease $c$ significantly.  Suppose one adds a point arbitrarily close to the
origin to $S'$; now, the expansion constant will be $|S'| / 2$.  So, although we
can bound the behavior of $c$ as $|S| \to \infty$ for $S$ from a stationary
distribution, we are not able to easily say much about its convergence behavior.

@@

\begin{lemma}
Consider a dataset $S$ with expansion constant $c$ and a subset $C \subseteq S$
such that every point in $C$ is separated by $\delta$.  Then, for any point
$p \not\in S$ and any radius $\rho \delta > 0$:
\begin{equation}
| B_S(p, \rho \delta) \cap C | \le c^{2 + \lceil \log_2 \rho \rceil}.
\end{equation}
\label{lem:packing}
\end{lemma}

>>>>

For simplicity, the algorithms considered in this paper will be presented in a
tree-independent context, as in \citet{curtin2013tree}, but the only type of
tree we will consider is the cover tree \citep{langford2006}, and the only type
of traversal we will consider is the cover tree pruning dual-tree traversal,
which we will describe later.

@@

As a consequence of this definition, if there exists a node $\mathscr{N}_i$,
containing the point $p_i$ at some scale $s_i$, then there will also exist a
self-child node $\mathscr{N}_{ic}$ containing the point $p_i$ at scale $s_i - 1$
which is a child of $\mathscr{N}_i$.  In addition, every descendant point of the
node $\mathscr{N}_i$ is contained within a ball of radius $2^{s_i + 1}$ centered
at the point $p_i$; therefore, $\lambda_i = 2^{s_i + 1}$ and $\mu_i = p_i$
(Table \ref{tab:notation}).

Note that the cover tree may be interpreted as an infinite-leveled tree, with
$C_{\infty}$ containing only the root point, $C_{-\infty} = S$, and all levels
between defined as above.  \citet{langford2006} find
this representation (which they call the {\it implicit} representation) easier
for description of their algorithms and some of their proofs.  But clearly,
this is not suitable for implementation; hence, there is an {\it explicit}
representation in which all nodes that have only a self-child are coalesced
upwards (that is, the node's self-child is removed, and the children of that
self-child are taken to be the children of the node).  Figure
\ref{fig:implicit-explicit} shows an example cover tree and highlights implicit
and explicit nodes.

In this work, we consider only the explicit representation of a cover tree, and
do not concern ourselves with the details of tree construction\footnote{A batch
construction algorithm is given by \citet{langford2006}, called
\texttt{Construct}.}.

@@

As a simple example, take $f(x)$ as a uniform spherical distribution in
$\mathcal{R}^d$: for any $|x| \le 1$, $f(x)$ is a constant; for $|x| > 1$, $f(x)
= 0$.  It is easy to see that $c_f$ in this situation is $2^d$, and thus for
some dataset $S$, $c$ must converge to that value as more and more points are
added to $S$.  Closed-form solutions for $c_f$ for more complex distributions
are less easy to derive; however, empirical speedup results from
\citet{langford2006} suggest the existence of datasets where $c$ is not strongly
dependent on $d$.  For instance, the \texttt{covtype} dataset has 54 dimensions
but the expansion constant is much smaller than other, lower-dimensional
datasets.
% To demonstrate this convergence, Table
%\ref{tab:ec_scaling} shows results for 10 trials of empirically calculated
%expansion constants for datasets drawn from this distribution, for varying $d$.

@@

There are some other important observations about the behavior of $c$.  Adding a
single point to $S$ may increase $c$ arbitrarily: consider a set $S$ distributed
entirely on the surface of a unit hypersphere.  If one adds a single point at
the origin, producing the set $S'$, then $c$ explodes to $|S'|$ whereas before
it may have been much smaller than $|S|$.  Adding a single point may also
decrease $c$ significantly.  Suppose one adds a point arbitrarily close to the
origin to $S'$; now, the expansion constant will be $|S'| / 2$.  Both of these
situations are degenerate cases not commonly encountered in real-world behavior;
we discuss them in order to point out that although we can bound the behavior of
$c$ as $|S| \to \infty$ for $S$ from a stationary distribution, we are not able
to easily say much about its convergence behavior.

@@

\begin{lemma}
Consider a dataset $S$ with expansion constant $c$ and a subset $C \subseteq S$
such that every point in $C$ is separated by $\delta$.  Then, for any point
$p$ (which may or may not be in $S$), and any radius $\rho \delta > 0$:
\begin{equation}
| B_S(p, \rho \delta) \cap C | \le c^{2 + \lceil \log_2 \rho \rceil}.
\end{equation}
\label{lem:packing}
\end{lemma}

### 2015-01-08 bounds-files/sec02-nearest-neighbor-search.tex
Minor improvements after a read-through.  I need to fix the bibliography (UGH) but this is all that I think is necessary in terms of changes to the paper.

The standard task of nearest neighbor search can be simply described: given a
query set $S_q$ and a reference set $S_r$, for each query point $p_q \in S_q$,
find the nearest neighbor $p_r$ in the reference set $S_r$.  The task is
well-studied and well-known, and there exist numerous approaches for both exact
and approximate nearest neighbor search, including the cover tree nearest
neighbor search algorithm due to \citet{langford2006}.  We will consider that
algorithm, but in a tree-independent sense as given by \citet{curtin2013tree};
this means that to describe the algorithm, we require only a \texttt{BaseCase()}
and \texttt{Score()} function; these are given in Algorithms
\ref{alg:nn_base_case} and \ref{alg:nn_score}, respectively.

Given a type of tree and traversal, these two functions store the current
nearest neighbor candidates in the array $N$ and their distances in the array
$D$. \citep[see][for a more complete
discussion of how this algorithm works and a proof of
correctness.]{curtin2013tree}  The
\texttt{Score()} function depends on a bound function $B(\mathscr{N}_q)$ which
represents the maximum distance that could possibly improve a nearest neighbor
candidate for any descendant point of the query node $\mathscr{N}_q$.  The
standard bound function $B(\mathscr{N}_q)$ used for cover trees is
%\begin{equation}
$B(\mathscr{N}_q) \le D[p_q] + 2^{s_q + 1}$ \citep[this is adapted from][]{langford2006}.
%\end{equation}

@@

\begin{thm}
Using cover trees, the standard cover tree pruning dual-tree traversal, and the
nearest neighbor search \texttt{BaseCase()} and \texttt{Score()} as given in
Algorithms \ref{alg:nn_base_case} and \ref{alg:nn_score}, respectively, and also
given a reference set $S_r$ with expansion constant $c_r$, a query set $S_q$,
the running time of the algorithm is bounded by $O(c_r^4 c_{qr}^5 (N +
i_t(\mathscr{T}_q) + \theta))$ with $i_t(\mathscr{T}_q$ and $\theta$ defined as
in Definition \ref{def:imbalance} and Lemma \ref{lem:extcase3}, respectively.
\label{thm:nns}
\end{thm}

@@

Assume that when $R^*$ is encountered, the maximum reference scale is
$s_r^{\max}$ and the query node is $\mathscr{N}_q$.  Every node $\mathscr{N}_r
\in R^*$ satisfies the property enforced in line
\ref{alg:line:ct-dual-ref-score} that
%\begin{equation}
$d_{\min}(\mathscr{N}_q, \mathscr{N}_r) \le B(\mathscr{N}_q)$.
%\end{equation}
%Relaxing this slightly,
%\begin{equation}
%d_{\min}(\mathscr{N}_q, \mathscr{N}_r) \le B_c(\mathscr{N}_q).
%\end{equation}
Using the definition of $d_{\min}(\cdot, \cdot)$ and $B(\cdot)$, we
expand the equation.  Note that $p_q$ is the point held in $\mathscr{N}_q$ and
$p_r$ is the point held in $\mathscr{N}_r$.  Also, take $\hat{p}_r$ to be the
current nearest neighbor candidate for $p_q$; that is, $D[p_q] = d(p_q,
\hat{p}_r)$.  Then,

\begin{eqnarray}
%d(p_q, p_r) - \lambda(\mathscr{N}_q) - \lambda(\mathscr{N}_r) &\le& D[p_q] +
%2^{s_q + 1} \\
%d(p_q, p_r) &\le& d(p_q, \hat{p}_r) + \lambda(\mathscr{N}_q) +
%\lambda(\mathscr{N}_r) + 2^{s_q + 1} \\
%d(p_q, p_r) - 2^{s_q + 1} - 2^{s_r + 1} &\le& D[p_q] + 2^{s_q + 1} \\
d(p_q, p_r) &\le& d(p_q, \hat{p}_r) + 2^{s_q + 1} + 2^{s_r + 1} + 2^{s_q + 1}
\label{eqn:pr_dist} \\
 &\le& d(p_q, \hat{p}_r) + 2(2^{s_r^{\max} + 1})
\end{eqnarray}

@@

\begin{equation}
P \subseteq B_{S_r}(p_q, d(p_q, \hat{p}_r) + 2(2^{s_r^{\max} + 1}))
\end{equation}

@@

\begin{eqnarray}
| B_{S_r \cup \{ p_q \}}(p_q, d(p_q, p_r^*) + 3(2^{s_r^{\max} + 1})) | &\le& |
B_{S_r \cup \{ p_q \}}(p_q, 4 d(p_q, p_r^*)) | \\
% &\le& | B_{S_{qr}}(p_q, 8 d(p_q, p_r^*)) | \\
% &\le& (c'_r)^3 | B_{S_r \cup \{ p_q \}}(p_q, d(p_q, p_r^*) / 2) | \\
 &\le& c_{qr}^3 | B_{S_r \cup \{ p_q \}}(p_q, d(p_q, p_r^*) / 2 |
\end{eqnarray}

@@

In the monochromatic case where $S_q = S_r$, the bound is $O(c^9 (N +
i_t(\mathscr{T}))$ because $c = c_r = c_{qr}$ and $\theta = 0$.  For
well-behaved trees where $i_t(\mathscr{T}_q)$ is linear or sublinear in $N$,
this represents the current best worse-case runtime bound for nearest neighbor
search.

>>>>

The standard task of nearest neighbor search can be simply described: given a
query set $S_q$ and a reference set $S_r$, for each query point $p_q \in S_q$,
find the nearest neighbor $p_r$ in the reference set $S_r$.  The task is
well-studied and well-known, and there exist numerous approaches for both exact
and approximate nearest neighbor search, including the cover tree nearest
neighbor search algorithm due to \citet{langford2006}.  We will consider that
algorithm, but in a tree-independent sense as given by \citet{curtin2013tree};
this means that to describe the algorithm, we require only a \texttt{BaseCase()}
and \texttt{Score()} function; these are given in Algorithms
\ref{alg:nn_base_case} and \ref{alg:nn_score}, respectively.  The
point-to-point \texttt{BaseCase()} function compares a query point $p_q$ and a
reference point $p_r$, updating the list of candidate neighbors for $p_q$ if
necessary.

The node-to-node \texttt{Score()} function determines if the entire
subtree of nodes under the reference node $\mathscr{N}_r$ can improve the
candidate neighbors for all descendant points of the query node $\mathscr{N}_q$;
if not, the node combination is pruned.  The \texttt{Score()} function depends
on the function $d_{\min}(\cdot, \cdot)$, which represents the minimum possible
distance between any two descendants of two nodes.  Its definition for cover
tree nodes is

\begin{equation}
d_{\min}(\mathscr{N}_q, \mathscr{N}_r) = d(p_q, p_r) - 2^{s_q + 1} - 2^{s_r +
1}.
\end{equation}

Given a type of tree and traversal, these two functions store the current
nearest neighbor candidates in the array $N$ and their distances in the array
$D$. \citep[See][for a more complete
discussion of how this algorithm works and a proof of
correctness.]{curtin2013tree}  The
\texttt{Score()} function depends on a bound function $B(\mathscr{N}_q)$ which
represents the maximum distance that could possibly improve a nearest neighbor
candidate for any descendant point of the query node $\mathscr{N}_q$.  The
standard bound function $B(\mathscr{N}_q)$ used for cover trees is adapted from
\citep{langford2006}:

\begin{equation}
B(\mathscr{N}_q) := D[p_q] + 2^{s_q + 1}
\end{equation}

@@

\begin{thm}
Using cover trees, the standard cover tree pruning dual-tree traversal, and the
nearest neighbor search \texttt{BaseCase()} and \texttt{Score()} as given in
Algorithms \ref{alg:nn_base_case} and \ref{alg:nn_score}, respectively, and also
given a reference set $S_r$ with expansion constant $c_r$, and a query set $S_q$,
the running time of the algorithm is bounded by $O(c_r^4 c_{qr}^5 (N +
i_t(\mathscr{T}_q) + \theta))$ with $i_t(\mathscr{T}_q$ and $\theta$ defined as
in Definition \ref{def:imbalance} and Lemma \ref{lem:extcase3}, respectively.
\label{thm:nns}
\end{thm}

@@

Assume that when $R^*$ is encountered, the maximum reference scale is
$s_r^{\max}$ and the query node is $\mathscr{N}_q$.  Every node $\mathscr{N}_r
\in R^*$ satisfies the property enforced in line
\ref{alg:line:ct-dual-ref-score} that
%\begin{equation}
$d_{\min}(\mathscr{N}_q, \mathscr{N}_r) \le B(\mathscr{N}_q)$.
%\end{equation}
%Relaxing this slightly,
%\begin{equation}
%d_{\min}(\mathscr{N}_q, \mathscr{N}_r) \le B_c(\mathscr{N}_q).
%\end{equation}
Using the definition of $d_{\min}(\cdot, \cdot)$ and $B(\cdot)$, we
expand the equation.  Note that $p_q$ is the point held in $\mathscr{N}_q$ and
$p_r$ is the point held in $\mathscr{N}_r$.  Also, take $\hat{p}_r$ to be the
current nearest neighbor candidate for $p_q$; that is, $D[p_q] = d(p_q,
\hat{p}_r)$ and $N[p_q] = \hat{p}_r$.  Then,

\begin{eqnarray}
%d(p_q, p_r) - \lambda(\mathscr{N}_q) - \lambda(\mathscr{N}_r) &\le& D[p_q] +
%2^{s_q + 1} \\
%d(p_q, p_r) &\le& d(p_q, \hat{p}_r) + \lambda(\mathscr{N}_q) +
%\lambda(\mathscr{N}_r) + 2^{s_q + 1} \\
%d(p_q, p_r) - 2^{s_q + 1} - 2^{s_r + 1} &\le& D[p_q] + 2^{s_q + 1} \\
d_{\min}(\mathscr{N}_q, \mathscr{N}_r) &\le& B(\mathscr{N}_q) \\
d(p_q, p_r) &\le& d(p_q, \hat{p}_r) + 2^{s_q + 1} + 2^{s_r + 1} + 2^{s_q + 1}
\label{eqn:pr_dist} \\
 &\le& d(p_q, \hat{p}_r) + 2(2^{s_r^{\max} + 1})
\end{eqnarray}

@@

\begin{equation}
P \subseteq B_{S_r}(p_q, d(p_q, \hat{p}_r) + 2(2^{s_r^{\max} + 1})).
\end{equation}

@@

\begin{eqnarray}
| B_{S_r \cup \{ p_q \}}(p_q, d(p_q, p_r^*) + 3(2^{s_r^{\max} + 1})) | &\le& |
B_{S_r \cup \{ p_q \}}(p_q, 4 d(p_q, p_r^*)) | \\
% &\le& | B_{S_{qr}}(p_q, 8 d(p_q, p_r^*)) | \\
% &\le& (c'_r)^3 | B_{S_r \cup \{ p_q \}}(p_q, d(p_q, p_r^*) / 2) | \\
 &\le& c_{qr}^3 | B_{S_r \cup \{ p_q \}}(p_q, d(p_q, p_r^*) / 2) |
\end{eqnarray}

@@

In the monochromatic case where $S_q = S_r$, the bound is $O(c^9 (N +
i_t(\mathscr{T}))$ because $c = c_r = c_{qr}$ and $\theta = 0$.  For
well-behaved trees where $i_t(\mathscr{T}_q)$ is linear or sublinear in $N$,
this represents the current tightest worst-case runtime bound for nearest
neighbor search.
